Assessing primitives performance\\ on multi-stage execution

### 2018-06-26 paper.tex
fix typos from review 1

High-level object-oriented programming languages are often implemented on top of a virtual machine (VM). As VMs has become more popular, different techniques have been set up using Just-In-Time (JIT) compilation to improve the runtime overall performance: method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.

@@

We implemented a first version of the optional primitive in Pharo/Smalltalk. We then implemented two versions compiling through C compilers to native code, using different parts of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions on strings of different sizes, showing that overall the most complex implementation is the fastest to execute.

@@

There is a performance overhead for the plugins compared to normal VM code. For example, accesses to object headers are memory representation dependent and therefore need to be done indirectly through calls to main VM executable, since the plugin cannot know ahead of time which memory representation is used. Such calls cannot be inlined by the C compiler\footnote{In the case of specific internal plugins, C compiler linking time optimizations can be enabled to limit this overhead.}.

@@

This difference is mainly due to the optimizations performed or not by both the compilers for the \legSlang version. As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar. On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in the figure).

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitive is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.

@@

To avoid paying the stack switch cost between the client Stack and the C stack, some VMs include an interpreter written on top of an abstract assembly. It allows the interpreter to have same register convention than the native code generated by the JIT, removing the stack switch overhead. For example, the V8 team recently wrote an interpreter named \emph{Ignition}. This interpreter is compiled ahead of time through their JIT back-end to native code. With their primitives and their interpreter written in such way, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives). In Javascript Core, a similar approach is taken with the interpreter named \emph{LLInt}~\cite{LLIntJSC}. LLInt is written in an abstract assembly code compiled ahead of time to native code compatible with the JIT's calling conventions.

@@

In this paper we compared four different implementations of the string primitives on top of the Cog VM. We showed the performance difference between each implementation for different string sizes. Choosing the right implementation for one's VM is a trade-off between maintenance cost and execution time, the quickest to execute the primitive is, the harder it is to maintain.

>>>>

High-level object-oriented programming languages are often implemented on top of a virtual machine (VM). As VMs have become more popular, different techniques have been set up using Just-In-Time (JIT) compilation to improve the overall runtime performance: method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.

@@

We implemented a first version of the optional primitive in Pharo. We then implemented two versions compiling through C compilers to native code, using different parts of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions on strings of different sizes, showing that overall the most complex implementation is the fastest to execute.

@@

There is a performance overhead for the plugins compared to normal VM code. For example, accesses to object headers are dependent on the memory representation and therefore need to be done indirectly through calls to main VM executable, since the plugin cannot know ahead of time which memory representation is used. Such calls cannot be inlined by the C compiler\footnote{In the case of specific internal plugins, C compiler linking time optimizations can be enabled to limit this overhead.}.

@@

This difference is mainly due to the optimizations performed or not by both the compilers for the \legSlang version. As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar. On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in Figure \ref{fig:Asm}).

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitives is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.

@@

To avoid paying the stack switch cost between the client Stack and the C stack, some VMs include an interpreter written on top of an abstract assembly. It allows the interpreter to have the same register convention as the native code generated by the JIT, removing the stack switch overhead. For example, the V8 team recently wrote an interpreter named \emph{Ignition}. This interpreter is compiled ahead of time through their JIT back-end to native code. With their primitives and their interpreter written in such way, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives). In Javascript Core, a similar approach is taken with the interpreter named \emph{LLInt}~\cite{LLIntJSC}. LLInt is written in an abstract assembly code compiled ahead of time to native code compatible with the JIT's calling conventions.

@@

In this paper we compared four different implementations of the string comparison primitive on top of the Cog VM. We showed the performance difference between each implementation for different string sizes. Choosing the right implementation for one's VM is a trade-off between maintenance cost and execution time, the quickest to execute the primitive is, the harder it is to maintain.
